Targeted interrogation of mutational hotspots in rheumatoid arthritis

### 2024-03-07 scomatic/reports/targeted_mutation_calling.Rmd
developing targeted mutation calling

```{r load}
# read targeted nanoseq 
tnanoseq <-
  '/lustre/scratch125/casm/team268im/lh22/synovium_nanoseq/ex_t_dndsout/dnds_nano_muts.tsv' %>%
  readr::read_tsv()

# read genes of interest
target_genes <-
  '/lustre/scratch125/casm/team268im/at31/RA_som_mut/scomatic/data/driver_genes/Sanger_TERT-v4_TE-95148282_hg19_highstringencyfilter_buccal_gene_list.tsv' %>%
  readr::read_tsv()
```

```{r dndscv_and_liftover}
# run dndscv
dnds <-
  dndscv::dndscv(
    unique(tnanoseq[, 1:5]),
    max_muts_per_gene_per_sample = Inf,
    max_coding_muts_per_sample = Inf,
    gene_list = target_genes$gene,
    outmats = T)

# run sitednds
sitednds <-
  dndscv::sitednds(dnds)

# get the significant genes
signif_genes <-
  dnds$sel_cv %>%
  tibble::as_tibble() %>%
  dplyr::filter(qglobal_cv < 0.1)

# get the sites with positively selected mutations
recursites <-
  sitednds$recursites %>%
  tibble::as_tibble()

# convert to granges (1-based!)
gr <- GRanges(
  seqnames = paste0('chr', recursites$chr),
  ranges = IRanges(start = recursites$pos, end = recursites$pos),
  gene = recursites$gene,
  pos_GRCh37 = recursites$pos)

# get chain file
chain <-
  rtracklayer::import.chain('data/liftOver/hg19ToHg38.over.chain')

# lift over
lifted <-
  rtracklayer::liftOver(gr, chain) %>%
  unlist() %>%
  tibble::as_tibble() %>%
  dplyr::transmute(chr = gsub('chr', '', seqnames), 
                   pos = start, gene, pos_GRCh37)

# update positions
recursites <-
  recursites %>%
  dplyr::rename(pos_GRCh37 = pos) %>%
  dplyr::left_join(lifted, by = c('chr', 'gene', 'pos_GRCh37')) %>%
  dplyr::select(chr, pos, gene, ref, mut, pos_GRCh37, everything())
```

# Pileup of BAMs from snRNA-seq data

```{r pileup}
# bam2r filtering arguments
min_phred <- 30
sam_flag <- 3844
mapq <- 25
nucleotides <- c("A", "T", "C", "G", "-", "INS", "a", "t", "c", "g", "_", "ins")

# list of id - ct - bam
bams <-
  mappings %>%
  {split(., .$id)} %>%
  purrr::map(function(df) {split(df$bam, df$celltype)})

# get bases at sites
sites <- xfun::cache_rds({

  sites <-
    bams %>% 
    purrr::map(function(id) {
      id %>% 
        purrr::map(function(ct_bam) {

          recursites %>%
            # dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*') %>% 
            purrr::pmap(function(chr, pos, ref, mut, gene, ...) {
              # testing: # recursites %>% dplyr::filter(gene == 'TNFRSF14', aachange == 'W12*') %>% as.list() %>% list2env(envir = globalenv())

              # positions
              positions <- (pos - 20):(pos + 20)

              # get reads at each position of the window
              ct_bam %>%
                deepSNV::bam2R(
                  chr = chr,
                  # extend pileup window 20bp either side of mutated position
                  start = pos - 20,
                  stop = pos + 20,
                  # quality controls
                  q = min_phred,
                  mask = sam_flag,
                  mq = mapq) %>%
                tibble::as_tibble() %>%
                dplyr::mutate(chr = chr, pos = positions, mut_pos = pos, gene = gene) %>%
                dplyr::select(chr, pos, mut_pos, gene, dplyr::all_of(nucleotides))

            }) %>%
            dplyr::bind_rows() %>%
            dplyr::rowwise() %>%
            dplyr::mutate(
              total = sum(dplyr::across(dplyr::all_of(nucleotides)))) %>%
            dplyr::relocate(dplyr::all_of(nucleotides), .after = dplyr::last_col())

        }) %>% dplyr::bind_rows(.id = 'celltype')
    }) %>% dplyr::bind_rows(.id = 'id')

  # return
  sites

}, file = 'sites.rds', rerun = params$rerun)

# add refs, compare
muts <-
  sites %>%
  dplyr::left_join(refs) %>%
  tidyr::pivot_longer(cols = dplyr::all_of(nucleotides),
                      values_to = 'nt_n_reads', names_to = 'nt') %>%
  # get number of ref and non-ref alleles at each position
  dplyr::mutate(n_ref = ifelse(ref == toupper(nt), nt_n_reads, 0),
                n_non_ref = ifelse(ref != toupper(nt), nt_n_reads, 0)) %>%
  dplyr::group_by(id, celltype, chr, pos, mut_pos, gene, total, ref) %>%
  dplyr::summarise(n_ref = sum(n_ref), n_non_ref = sum(n_non_ref)) %>%
  dplyr::mutate(vaf = n_non_ref / total)
```

>>>>

```{r load_files}
list.files(pattern = '_reads.tsv') %>%
  purrr::set_names(., gsub('_reads.tsv', '', .)) %>%
  purrr::map(function(file) {
    reads <- readr::read_tsv(reads)
  })
```
